---
title: "Linear regression and logistic regression with missing covariates"
author: "Wei Jiang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{misaem tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction of misaem

`misaem` is a package to perform linear regression and logistic regression with missing data, under MCAR (Missing completely at random) and MAR (Missing at random) mechanisms. The covariates are assumed to be continuous variables. The methodology implemented is based on maximization of the observed likelihood using EM-types of algorithms. The package includes:

1. Parameters estimation:
* for linear regression, we consider a joint Gaussian distribution for covariates and response, then the `norm` package allows to estimate the mean vector and a variance covariance matrix with the EM algorithm and SWEEP operator. Finally we have reshaped them to obtain the regression coefficient.
* for logistic regression, with a stochastic approximation version of EM algorithm (SAEM) based on Metropolis-Hasting sampling.
2. Estimation of standard deviation for estimated parameters:
* for linear regression, use the property that the Gram matrix of random variables (estimates of regression coefficients) approximates their covariance matrix;
* for logistic regression, with Louis formula.
3. Model selection procedure based on BIC. 
```{r}
library(misaem)
```


## Linear regression
### Synthetic dataset
Let's generate a synthetic example of classical linear regression. We first generate a design matrix of size $n = 50$ times $p = 2$ by drawing each observation from a multivariate normal distribution $\mathcal{N}(\mu, \Sigma)$. We consider as the true values for the parameters:
\begin{equation*}
\begin{split}
\mu &= (1, 1),\\
\Sigma & = \begin{bmatrix}
1 & 1\\
1 & 4\\
\end{bmatrix}
\end{split}
\end{equation*}
Then, we generate the response according to the linear regression model with coefficient $\beta = (2, 3, -1)$ and variance of noise vector $\sigma^2 = 0.25$.
```{r}
set.seed(1)
n <- 50 # number of rows
p <- 2 # number of explanatory variables

# Generate complete design matrix
library(MASS)
mu.X <- c(1, 1) 
Sigma.X <- matrix(c(1, 1, 1, 4), nrow = 2)
X.complete <- mvrnorm(n, mu.X, Sigma.X)

# Generate response
b <- c(2, 3, -1) # regression coefficient
sigma.eps <- 0.25 # noise variance
y <- cbind(rep(1, n), X.complete) %*% b + rnorm(n, 0, sigma.eps)
```
Then we randomly introduced 15\% of missing values in the covariates according to the MCAR (Missing completely at random) mechanism. To do so, we use the function `ampute` from the R package `mice`. For more details about how to generate missing values of different mechanisms, see the resource website of missing values [Rmisstastic](https://rmisstastic.netlify.app/).
```{r}
library(mice)
# Add missing values
yX.miss <- ampute(data.frame(y, X.complete), 0.15, patterns = matrix(
  c(0, 0, 1, 0, 1, 0, 1, 0, 0, 0, 1, 1, 1, 0, 1, 1, 1, 0), 
  ncol = 3, byrow = TRUE), freq = c(1, 1, 1, 2, 2, 2) / 9, 
  mech = "MCAR", bycases = FALSE)
y.obs <- yX.miss$amp[, 1]              # responses
X.obs <- as.matrix(yX.miss$amp[, 2:3]) # covariates with NAs
```
Have a look at our synthetic dataset:
```{r}
head(X.obs)
```
Check the percentage of missing values:
```{r}
sum(is.na(X.obs))/(n*p) 
```

### Estimation for linear regression with missing values
The main function in our package to fit linear regression with missingness is `miss.lm` function. The function `miss.lm` mimics the structure of widely used function `lm` for the case without missing values. It takes an object of class `formula` (a symbolic description of the model to be fitted) and the data frame as the input. Here we apply this function with its default options.

```{r}
# Estimate regression using EM with NA 
df.obs = data.frame(y, X.obs)
miss.list = miss.lm(y~., data = df.obs)
```
Then it returns an object of self-defined class `miss.lm`, which consists of the estimation of parameters, their standard error and observed log-likelihood. We can print or summarize the obtained results as follows:
```{r}
print(miss.list)
print(summary(miss.list)) 
summary(miss.list)$coef 
```
Self-defined parameters can be also taken such as the maximum number of iterations (`maxruns`), the convergence tolerance (`tol_em`) and the logical indicating if the iterations should be reported (`print_iter`). 
```{r}
# Estimate regression using self-defined parameters
miss.list2 = miss.lm(y~., data = df.obs, print_iter = FALSE, maxruns = 500, tol_em = 1e-4)
print(miss.list2)
```

### Model selection
The function `miss.lm.model.select` adapts a BIC criterion and step-wise method to return the best model selected. We add a null variable with missing values to check if the function can distinguish it from the true variables.
```{r}
# Add null variable with NA
X.null <- mvrnorm(n, 1, 1)
patterns <- runif(n)<0.15 # missing completely at random
X.null[patterns] <- NA
X.obs.null <- cbind.data.frame(X.obs, X.null)

# Without model selection
df.obs.null = data.frame(y, X.obs.null)
miss.list.null = miss.lm(y~., data = df.obs.null)
print(miss.list.null)

# Model selection
miss.model = miss.lm.model.select(y, X.obs.null)
print(miss.model)
```

### Prediction on test set
In order to evaluate the prediction performance, we generate a test set of size $nt = 20$ times $p = 2$ following the same distribution as the previous design matrix, and we add or not 15\% of missing values.
```{r}
# Prediction
# Generate dataset
set.seed(200)
nt <- 20  # number of new observations
Xt <- mvrnorm(nt, mu.X, Sigma.X)

# Add missing values
Xt.miss <- ampute(data.frame(Xt), 0.15, patterns = matrix(
  c(0, 1, 1, 0), 
  ncol = 2, byrow = TRUE), freq = c(1, 1) /2, 
  mech = "MCAR", bycases = FALSE)
Xt.obs <- as.matrix(Xt.miss$amp) # covariates with NAs
```

The prediction can be performed for a complete test set:
```{r}
#train with NA + test no NA
miss.comptest.pred = predict(miss.list2, data.frame(Xt), seed = 100)
print(miss.comptest.pred)
```

And we can also apply the function when both train set and test set have missing values:
```{r}
#both train & test with NA
miss.pred = predict(miss.list2, data.frame(Xt.obs), seed = 100)
print(miss.pred)
```

## Logistic regression
### Synthetic dataset
We first generate a design matrix of size $n=500$ times $p=5$ by drawing each observation from a multivariate normal distribution $\mathcal{N}(\mu, \Sigma)$. Then, we generate the response according to the logistic regression model.

We consider as the true values for the parameters
\begin{equation*}
\begin{split}
\beta &= (0, 1, -1, 1, 0, -1),\\
\mu &= (1,2,3,4,5),\\
\Sigma &= \text{diag}(\sigma)C \text{diag}(\sigma),
\end{split}
\end{equation*} 
where the $\sigma$ is the vector of standard deviations 
$$\sigma=(1,2,3,4,5)$$  
and  $C$ the correlation matrix 
$$C = \begin{bmatrix}
1  & 0.8 & 0 & 0 &   0\\
0.8 & 1 & 0 & 0  &  0\\
0  & 0 & 1 & 0.3 &   0.6\\
0 & 0 & 0.3 & 1 &  0.7\\
0 & 0 & 0.6 & 0.7 &  1\\
\end{bmatrix}.$$ 

```{r}
# Generate dataset
set.seed(200)
n <- 500  # number of subjects
p <- 5     # number of explanatory variables
mu.star <- 1:p  #rep(0,p)  # mean of the explanatory variables
sd <- 1:p # rep(1,p) # standard deviations
C <- matrix(c(   # correlation matrix
1,   0.8, 0,   0,   0,
0.8, 1,   0,   0,   0,
0,   0,   1,   0.3, 0.6,
0,   0,   0.3, 1,   0.7,
0,   0,   0.6, 0.7, 1), nrow=p)
Sigma.star <- diag(sd)%*%C%*%diag(sd) # covariance matrix
beta.star <- c(1, -1, 1, 1, -1) # coefficients
beta0.star <- 0  # intercept
beta.true = c(beta0.star,beta.star)

# Design matrix
X.complete <- matrix(rnorm(n*p), nrow=n)%*%chol(Sigma.star)+
              matrix(rep(mu.star,n), nrow=n, byrow = TRUE)

# Reponse vector
p1 <- 1/(1+exp(-X.complete%*%beta.star-beta0.star))
y <- as.numeric(runif(n)<p1)
```

Then we randomly introduced 10\% of missing values in the covariates according to the MCAR (Missing completely at random) mechanism.
```{r}
# Generate missingness
set.seed(200)
p.miss <- 0.10
patterns <- runif(n*p)<p.miss # missing completely at random
X.obs <- X.complete
X.obs[patterns] <- NA
```

Have a look at our synthetic dataset:
```{r}
head(X.obs)
```

### Estimation for logistic regression with missingness
The main function for fitting logistic regression with missing covariates in our package is `miss.glm` function, which mimics the structure of widely used function `glm`. Note that we don't need to specify the binomial family in the input of `miss.glm` function. Here we apply this function with its default options, and then we can print or summarize the obtained results as follows:
```{r}
df.obs = data.frame(y, X.obs)

#logistic regression with NA 
miss.list = miss.glm(y~., data = df.obs, seed = 100)
print(miss.list)
print(summary(miss.list)) 
summary(miss.list)$coef 
```

### Model selection 
To perform model selection with missing values, we adapt criterion BIC and step-wise method. The function `miss.glm.model.select` outputs the best model selected. With the current implementation, when $p$ is greater than 20, it may encounter computational difficulties for the BIC based model selection. In the following simulation, we add a null variable with missing values to check if the function can distinguish it from the true variables.

```{r}
# Add null variable with NA
X.null <- mvrnorm(n, 1, 1)
patterns <- runif(n)<0.10 # missing completely at random
X.null[patterns] <- NA
X.obs.null <- cbind.data.frame(X.obs, X.null)

# Without model selection
df.obs.null = data.frame(y, X.obs.null)
miss.list.null = miss.glm(y~., data = df.obs.null)
print(miss.list.null)

# model selection for SAEM
miss.model = miss.glm.model.select(y, X.obs.null)
print(miss.model)
```



### Prediction on test set
In order to evaluate the prediction performance, we generate a test set of size $nt = 100$ times $p = 5$ following the same distribution as the design matrix, and without and with 10\% of missing values. We  evaluate the prediction quality with a confusion matrix.
```{r}
# Generate test set with missingness
set.seed(200)
nt = 100
X.test <- matrix(rnorm(nt*p), nrow=nt)%*%chol(Sigma.star)+
          matrix(rep(mu.star,nt), nrow = nt, byrow = TRUE)

# Generate the test set
p1 <- 1/(1+exp(-X.test%*%beta.star-beta0.star))
y.test <- as.numeric(runif(nt)<p1)

# Generate missingness on test set
p.miss <- 0.10
X.test[runif(nt*p)<p.miss] <- NA

# Prediction on test set
pr.saem <- predict(miss.list, data.frame(X.test))

# Confusion matrix
pred.saem = (pr.saem>0.5)*1
table(y.test,pred.saem )
```
 
## Reference 
Logistic Regression with Missing Covariates
-- Parameter Estimation, Model Selection
and Prediction (2020, Jiang W., Josse J., Lavielle M., TraumaBase Group), [Computational Statistics \& Data Analysis](https://doi.org/10.1016/j.csda.2019.106907).
